Capstone Movielens Report

### 2025-05-07 reports/capstone-movielens-report.draft4.Rmd
Draft report updated.

### Defining Logging and Mesuring helper functions
\

@@

::: {.noteblock data-latex=""}
The full source code of these functions is available in the [Logging Helper function section of the Capstone MovieLens Main R Script](https://github.com/AzKurban-edX-DS/Capstone-MovieLens/blob/2c5d96ac98a8b8fa4151aa8295f91ac29d9472b3/r/src/capstone-movielens.main.R#L1).
:::

@@

::: {.noteblock data-latex=""}
The full source code of the function listed above is available in the [`Initialize input datasets` section of the `data.helper.functions.R` script on _GitHub_](https://github.com/AzKurban-edX-DS/Capstone-MovieLens/blob/2c5d96ac98a8b8fa4151aa8295f91ac29d9472b3/r/src/support-functions/data.helper.functions.R#L86)
:::

@@

::: {.noteblock data-latex=""}
The source code of the function mentioned above is also available in the [`Initialize input datasets` section of the `data.helper.functions.R` script on _GitHub_](https://github.com/AzKurban-edX-DS/Capstone-MovieLens/blob/2c5d96ac98a8b8fa4151aa8295f91ac29d9472b3/r/src/support-functions/data.helper.functions.R#L86)
:::

@@

::: {.noteblock data-latex=""}
This code snippet is a part of the [`make_source_datasets` _function_ code](https://github.com/AzKurban-edX-DS/Capstone-MovieLens/blob/2c5d96ac98a8b8fa4151aa8295f91ac29d9472b3/r/src/support-functions/data.helper.functions.R#L87) described above.
:::

@@

::: {.noteblock data-latex=""}
The [`sample_train_validation_sets` function](https://github.com/AzKurban-edX-DS/Capstone-MovieLens/blob/2c5d96ac98a8b8fa4151aa8295f91ac29d9472b3/r/src/support-functions/data.helper.functions.R#L15) is defined in the same script as the `make_source_datasets` one, from where it is called.  
:::

@@

For our further analysis, we are going to use the following _common helper functions:

@@

@@

mu.RMSE <- sqrt(mean(mu.MSEs))
mu.RMSE
```
::: {.noteblock data-latex=""}
For the _Mean Squared Error_ data visualization we used [`data.plot` function](https://github.com/AzKurban-edX-DS/Capstone-MovieLens/blob/914dd47023dc21b6b06a0ca412b7f7281602275a/r/src/support-functions/common-helper.functions.R#L630) defined in [_Model Data Visualization_ section](https://github.com/AzKurban-edX-DS/Capstone-MovieLens/blob/914dd47023dc21b6b06a0ca412b7f7281602275a/r/src/support-functions/common-helper.functions.R#L484) of the [common-helper.functions.R](https://github.com/AzKurban-edX-DS/Capstone-MovieLens/blob/main/r/src/support-functions/common-helper.functions.R) script.
:::

Here we also used [`naive_model_MSEs` function](https://github.com/AzKurban-edX-DS/Capstone-MovieLens/blob/914dd47023dc21b6b06a0ca412b7f7281602275a/r/src/support-functions/common-helper.functions.R#L54) defined in the [same script](https://github.com/AzKurban-edX-DS/Capstone-MovieLens/blob/main/r/src/support-functions/common-helper.functions.R) to compute _Mean Squared Errors_ using _5-Fold Cross Validation_ method:

@@

One more function we will need for further analysis of the current model is the [`naive_model_RMSE` one](https://github.com/AzKurban-edX-DS/Capstone-MovieLens/blob/914dd47023dc21b6b06a0ca412b7f7281602275a/r/src/support-functions/common-helper.functions.R#L59) defined in the [common-helper.functions.R](https://github.com/AzKurban-edX-DS/Capstone-MovieLens/blob/main/r/src/support-functions/common-helper.functions.R) script already mentioned above:
```{r eval=FALSE }
naive_model_RMSE <- function(val){
  sqrt(mean(naive_model_MSEs(val)))
}
```

@@

### Taking into account User effects
\
To improve our model let's now take into consideration user effects as explained in [Section _23.4 User effects_ of the *Course Textbook*](https://rafalab.dfci.harvard.edu/dsbook-part-2/highdim/regularization.html#user-effects). 
If we visualize the average rating for each user the way the [the author](https://x.com/rafalab) shows, we can see that there is substantial variability in the average ratings across users: 
```{r}
hist(edx.user_mean_ratings$mean_rating, nclass = 30)
print("A histogram of the User Mean Rating distribution has been plotted.")
```

>>>>

### Defining Logging and Time Measuring Helper Functions
\

@@

::: {.noteblock data-latex=""}
The full source code of these functions is available in the [Logging Helper function section](https://github.com/AzKurban-edX-DS/Capstone-MovieLens/blob/main/r/src/capstone-movielens.main.R#L20) of the [Capstone MovieLens Main R Script](https://github.com/AzKurban-edX-DS/Capstone-MovieLens/blob/main/r/src/capstone-movielens.main.R).
:::

@@

::: {.noteblock data-latex=""}
The full source code of the function listed above is available in the [Initialize input datasets](https://github.com/AzKurban-edX-DS/Capstone-MovieLens/blob/main/r/src/support-functions/data.helper.functions.R#L86) section of the [data.helper.functions.R](https://github.com/AzKurban-edX-DS/Capstone-MovieLens/blob/main/r/src/support-functions/data.helper.functions.R) script on _GitHub_.
:::

@@

::: {.noteblock data-latex=""}
The source code of the function mentioned above is also available in the [Initialize input datasets](https://github.com/AzKurban-edX-DS/Capstone-MovieLens/blob/main/r/src/support-functions/data.helper.functions.R#L86) section of the [data.helper.functions.R](https://github.com/AzKurban-edX-DS/Capstone-MovieLens/blob/main/r/src/support-functions/data.helper.functions.R) script on _GitHub_.
:::

@@

::: {.noteblock data-latex=""}
This code snippet is a part of the [make_source_datasets](https://github.com/AzKurban-edX-DS/Capstone-MovieLens/blob/main/r/src/support-functions/data.helper.functions.R#L87) _function_ code described above.
:::

@@

::: {.noteblock data-latex=""}
The [sample_train_validation_sets](https://github.com/AzKurban-edX-DS/Capstone-MovieLens/blob/main/r/src/support-functions/data.helper.functions.R#L15) function is defined in the same script as the [make_source_datasets](https://github.com/AzKurban-edX-DS/Capstone-MovieLens/blob/main/r/src/support-functions/data.helper.functions.R#L87)` one, from where it is called.
:::

@@

For our further analysis, we are going to use the following _common helper functions_:

@@

::: {.noteblock data-latex=""}
All the _common helper functions_, including those described above, are defined in the [common-helper.functions.R](https://github.com/AzKurban-edX-DS/Capstone-MovieLens/blob/main/r/src/support-functions/common-helper.functions.R) script on _GitHub_. 
:::

@@

mu.RMSE <- sqrt(mean(mu.MSEs))
mu.RMSE
```
::: {.noteblock data-latex=""}
For the _Mean Squared Error_ data visualization we used [data.plot](https://github.com/AzKurban-edX-DS/Capstone-MovieLens/blob/main/r/src/support-functions/data.helper.functions.R#L592) function] defined in the [Data Visualization](https://github.com/AzKurban-edX-DS/Capstone-MovieLens/blob/main/r/src/support-functions/data.helper.functions.R#L447) section of the [data.helper.function.R](https://github.com/AzKurban-edX-DS/Capstone-MovieLens/blob/main/r/src/support-functions/data.helper.functions.R) script.
:::

```{r eval=FALSE}
data.plot <- function(data, 
                      title, 
                      xname, 
                      yname, 
                      xlabel = NULL, 
                      ylabel = NULL,
                      line_col = "blue",
                      # scale = 1,
                      normalize = FALSE) {
  y <- data[, yname]

  if (normalize) {
    y <- y - min(y)
  }

  if (is.null(xlabel)) {
    xlabel = xname
  }
  if (is.null(ylabel)) {
    ylabel = yname
  }

  aes_mapping <- aes(x = data[, xname], y = y)

  data |> 
    ggplot(mapping = aes_mapping) +
    ggtitle(title) +
    xlab(xlabel) +
    ylab(ylabel) +
    geom_point() + 
    geom_line(color=line_col)
}
```

Here we also used [naive_model_MSEs](https://github.com/AzKurban-edX-DS/Capstone-MovieLens/blob/main/r/src/support-functions/common-helper.functions.R#L54) function defined in the [common-helper.functions.R](https://github.com/AzKurban-edX-DS/Capstone-MovieLens/blob/main/r/src/support-functions/common-helper.functions.R) script (already mentioned above) to compute _Mean Squared Errors_ using _5-Fold Cross Validation_ method:

@@

One more function, defined in the [same script](https://github.com/AzKurban-edX-DS/Capstone-MovieLens/blob/main/r/src/support-functions/common-helper.functions.R), that we will need for further analysis of the current model, is the [naive_model_RMSE](https://github.com/AzKurban-edX-DS/Capstone-MovieLens/blob/main/r/src/support-functions/common-helper.functions.R#L59) one:
```{r eval=FALSE }
naive_model_RMSE <- function(val){
  sqrt(mean(naive_model_MSEs(val)))
}
```

@@

### Taking into account User Effect
\
To improve our model let's now take into consideration user effects as explained in [Section _23.4 User effects_ of the *Course Textbook*](https://rafalab.dfci.harvard.edu/dsbook-part-2/highdim/regularization.html#user-effects). 
If we visualize the average rating for each user the way the [the author](https://x.com/rafalab) shows, we can see that there is substantial variability in the average ratings across users: 
```{r}
hist(edx.user_mean_ratings$mean_rating, nclass = 30)
print("A histogram of the User Mean Rating distribution has been plotted.")
```

### 2025-05-07 reports/capstone-movielens-report.draft4.tex
Draft report updated.

\title{Capstone Movielens Report}
\author{Azamat Kurbanaev}
\date{2025-05-07}

@@

\includegraphics{capstone-movielens-report.draft4_files/figure-latex/unnamed-chunk-22-1.pdf}

@@

\includegraphics{capstone-movielens-report.draft4_files/figure-latex/unnamed-chunk-25-1.pdf}

@@

\includegraphics{capstone-movielens-report.draft4_files/figure-latex/unnamed-chunk-26-1.pdf}

@@

\includegraphics{capstone-movielens-report.draft4_files/figure-latex/unnamed-chunk-29-1.pdf}

@@

\includegraphics{capstone-movielens-report.draft4_files/figure-latex/unnamed-chunk-30-1.pdf}

@@

\begin{noteblock}
The full source code of these functions is available in the
\href{https://github.com/AzKurban-edX-DS/Capstone-MovieLens/blob/2c5d96ac98a8b8fa4151aa8295f91ac29d9472b3/r/src/capstone-movielens.main.R\#L1}{Logging
Helper function section of the Capstone MovieLens Main R Script}.

@@

\begin{noteblock}
The full source code of the function listed above is available in the
\href{https://github.com/AzKurban-edX-DS/Capstone-MovieLens/blob/2c5d96ac98a8b8fa4151aa8295f91ac29d9472b3/r/src/support-functions/data.helper.functions.R\#L86}{\texttt{Initialize\ input\ datasets}
section of the \texttt{data.helper.functions.R} script on \emph{GitHub}}

@@

\begin{noteblock}
The source code of the function mentioned above is also available in the
\href{https://github.com/AzKurban-edX-DS/Capstone-MovieLens/blob/2c5d96ac98a8b8fa4151aa8295f91ac29d9472b3/r/src/support-functions/data.helper.functions.R\#L86}{\texttt{Initialize\ input\ datasets}
section of the \texttt{data.helper.functions.R} script on \emph{GitHub}}

@@

\begin{noteblock}
This code snippet is a part of the
\href{https://github.com/AzKurban-edX-DS/Capstone-MovieLens/blob/2c5d96ac98a8b8fa4151aa8295f91ac29d9472b3/r/src/support-functions/data.helper.functions.R\#L87}{\texttt{make\_source\_datasets}
\emph{function} code} described above.

@@

\begin{noteblock}
The
\href{https://github.com/AzKurban-edX-DS/Capstone-MovieLens/blob/2c5d96ac98a8b8fa4151aa8295f91ac29d9472b3/r/src/support-functions/data.helper.functions.R\#L15}{\texttt{sample\_train\_validation\_sets}
function} is defined in the same script as the
\texttt{make\_source\_datasets} one, from where it is called.

@@

For our further analysis, we are going to use the following \_common
helper functions:

@@

@@

\includegraphics{capstone-movielens-report.draft4_files/figure-latex/unnamed-chunk-49-1.pdf}

@@

\begin{noteblock}
For the \emph{Mean Squared Error} data visualization we used
\href{https://github.com/AzKurban-edX-DS/Capstone-MovieLens/blob/914dd47023dc21b6b06a0ca412b7f7281602275a/r/src/support-functions/common-helper.functions.R\#L630}{\texttt{data.plot}
function} defined in
\href{https://github.com/AzKurban-edX-DS/Capstone-MovieLens/blob/914dd47023dc21b6b06a0ca412b7f7281602275a/r/src/support-functions/common-helper.functions.R\#L484}{\emph{Model
Data Visualization} section} of the
\href{https://github.com/AzKurban-edX-DS/Capstone-MovieLens/blob/main/r/src/support-functions/common-helper.functions.R}{common-helper.functions.R}
script.

@@

Here we also used
\href{https://github.com/AzKurban-edX-DS/Capstone-MovieLens/blob/914dd47023dc21b6b06a0ca412b7f7281602275a/r/src/support-functions/common-helper.functions.R\#L54}{\texttt{naive\_model\_MSEs}
function} defined in the
\href{https://github.com/AzKurban-edX-DS/Capstone-MovieLens/blob/main/r/src/support-functions/common-helper.functions.R}{same
script} to compute \emph{Mean Squared Errors} using \emph{5-Fold Cross
Validation} method:

@@

One more function we will need for further analysis of the current model
is the
\href{https://github.com/AzKurban-edX-DS/Capstone-MovieLens/blob/914dd47023dc21b6b06a0ca412b7f7281602275a/r/src/support-functions/common-helper.functions.R\#L59}{\texttt{naive\_model\_RMSE}
one} defined in the
\href{https://github.com/AzKurban-edX-DS/Capstone-MovieLens/blob/main/r/src/support-functions/common-helper.functions.R}{common-helper.functions.R}
script already mentioned above:

@@

\includegraphics{capstone-movielens-report.draft4_files/figure-latex/unnamed-chunk-53-1.pdf}

@@

\subsubsection{Taking into account User
effects}\label{taking-into-account-user-effects}

@@

\includegraphics{capstone-movielens-report.draft4_files/figure-latex/unnamed-chunk-54-1.pdf}

@@

\includegraphics{capstone-movielens-report.draft4_files/figure-latex/unnamed-chunk-66-1.pdf}

>>>>

\title{Capstone Movielens Report}
\author{Azamat Kurbanaev}
\date{2025-05-08}

@@

\includegraphics{capstone-movielens-report.draft4_files/figure-latex/unnamed-chunk-11-1.pdf}

@@

\includegraphics{capstone-movielens-report.draft4_files/figure-latex/unnamed-chunk-14-1.pdf}

@@

\includegraphics{capstone-movielens-report.draft4_files/figure-latex/unnamed-chunk-15-1.pdf}

@@

\includegraphics{capstone-movielens-report.draft4_files/figure-latex/unnamed-chunk-18-1.pdf}

@@

\includegraphics{capstone-movielens-report.draft4_files/figure-latex/unnamed-chunk-19-1.pdf}

@@

\begin{noteblock}
The full source code of these functions is available in the
\href{https://github.com/AzKurban-edX-DS/Capstone-MovieLens/blob/main/r/src/capstone-movielens.main.R\#L20}{Logging
Helper function section} of the
\href{https://github.com/AzKurban-edX-DS/Capstone-MovieLens/blob/main/r/src/capstone-movielens.main.R}{Capstone
MovieLens Main R Script}.

@@

\begin{noteblock}
The full source code of the function listed above is available in the
\href{https://github.com/AzKurban-edX-DS/Capstone-MovieLens/blob/main/r/src/support-functions/data.helper.functions.R\#L86}{Initialize
input datasets} section of the
\href{https://github.com/AzKurban-edX-DS/Capstone-MovieLens/blob/main/r/src/support-functions/data.helper.functions.R}{data.helper.functions.R}
script on \emph{GitHub}.

@@

\begin{noteblock}
The source code of the function mentioned above is also available in the
\href{https://github.com/AzKurban-edX-DS/Capstone-MovieLens/blob/main/r/src/support-functions/data.helper.functions.R\#L86}{Initialize
input datasets} section of the
\href{https://github.com/AzKurban-edX-DS/Capstone-MovieLens/blob/main/r/src/support-functions/data.helper.functions.R}{data.helper.functions.R}
script on \emph{GitHub}.

@@

\begin{noteblock}
This code snippet is a part of the
\href{https://github.com/AzKurban-edX-DS/Capstone-MovieLens/blob/main/r/src/support-functions/data.helper.functions.R\#L87}{make\_source\_datasets}
\emph{function} code described above.

@@

\begin{noteblock}
The
\href{https://github.com/AzKurban-edX-DS/Capstone-MovieLens/blob/main/r/src/support-functions/data.helper.functions.R\#L15}{sample\_train\_validation\_sets}
function is defined in the same script as the
\href{https://github.com/AzKurban-edX-DS/Capstone-MovieLens/blob/main/r/src/support-functions/data.helper.functions.R\#L87}{make\_source\_datasets}`
one, from where it is called.

@@

For our further analysis, we are going to use the following \emph{common
helper functions}:

@@

\begin{noteblock}
All the \emph{common helper functions}, including those described above,
are defined in the
\href{https://github.com/AzKurban-edX-DS/Capstone-MovieLens/blob/main/r/src/support-functions/common-helper.functions.R}{common-helper.functions.R}
script on \emph{GitHub}.

@@

\includegraphics{capstone-movielens-report.draft4_files/figure-latex/unnamed-chunk-38-1.pdf}

@@

\begin{noteblock}
For the \emph{Mean Squared Error} data visualization we used
\href{https://github.com/AzKurban-edX-DS/Capstone-MovieLens/blob/main/r/src/support-functions/data.helper.functions.R\#L592}{data.plot}
function{]} defined in the
\href{https://github.com/AzKurban-edX-DS/Capstone-MovieLens/blob/main/r/src/support-functions/data.helper.functions.R\#L447}{Data
Visualization} section of the
\href{https://github.com/AzKurban-edX-DS/Capstone-MovieLens/blob/main/r/src/support-functions/data.helper.functions.R}{data.helper.function.R}
script.

@@

\begin{Shaded}
\begin{Highlighting}[]
\NormalTok{data.plot }\OtherTok{\textless{}{-}} \ControlFlowTok{function}\NormalTok{(data, }
\NormalTok{                      title, }
\NormalTok{                      xname, }
\NormalTok{                      yname, }
                      \AttributeTok{xlabel =} \ConstantTok{NULL}\NormalTok{, }
                      \AttributeTok{ylabel =} \ConstantTok{NULL}\NormalTok{,}
                      \AttributeTok{line\_col =} \StringTok{"blue"}\NormalTok{,}
                      \CommentTok{\# scale = 1,}
                      \AttributeTok{normalize =} \ConstantTok{FALSE}\NormalTok{) \{}
\NormalTok{  y }\OtherTok{\textless{}{-}}\NormalTok{ data[, yname]}

  \ControlFlowTok{if}\NormalTok{ (normalize) \{}
\NormalTok{    y }\OtherTok{\textless{}{-}}\NormalTok{ y }\SpecialCharTok{{-}} \FunctionTok{min}\NormalTok{(y)}
\NormalTok{  \}}

  \ControlFlowTok{if}\NormalTok{ (}\FunctionTok{is.null}\NormalTok{(xlabel)) \{}
\NormalTok{    xlabel }\OtherTok{=}\NormalTok{ xname}
\NormalTok{  \}}
  \ControlFlowTok{if}\NormalTok{ (}\FunctionTok{is.null}\NormalTok{(ylabel)) \{}
\NormalTok{    ylabel }\OtherTok{=}\NormalTok{ yname}
\NormalTok{  \}}

\NormalTok{  aes\_mapping }\OtherTok{\textless{}{-}} \FunctionTok{aes}\NormalTok{(}\AttributeTok{x =}\NormalTok{ data[, xname], }\AttributeTok{y =}\NormalTok{ y)}

\NormalTok{  data }\SpecialCharTok{|\textgreater{}} 
    \FunctionTok{ggplot}\NormalTok{(}\AttributeTok{mapping =}\NormalTok{ aes\_mapping) }\SpecialCharTok{+}
    \FunctionTok{ggtitle}\NormalTok{(title) }\SpecialCharTok{+}
    \FunctionTok{xlab}\NormalTok{(xlabel) }\SpecialCharTok{+}
    \FunctionTok{ylab}\NormalTok{(ylabel) }\SpecialCharTok{+}
    \FunctionTok{geom\_point}\NormalTok{() }\SpecialCharTok{+} 
    \FunctionTok{geom\_line}\NormalTok{(}\AttributeTok{color=}\NormalTok{line\_col)}
\NormalTok{\}}
\end{Highlighting}
\end{Shaded}

Here we also used
\href{https://github.com/AzKurban-edX-DS/Capstone-MovieLens/blob/main/r/src/support-functions/common-helper.functions.R\#L54}{naive\_model\_MSEs}
function defined in the
\href{https://github.com/AzKurban-edX-DS/Capstone-MovieLens/blob/main/r/src/support-functions/common-helper.functions.R}{common-helper.functions.R}
script (already mentioned above) to compute \emph{Mean Squared Errors}
using \emph{5-Fold Cross Validation} method:

@@

One more function, defined in the
\href{https://github.com/AzKurban-edX-DS/Capstone-MovieLens/blob/main/r/src/support-functions/common-helper.functions.R}{same
script}, that we will need for further analysis of the current model, is
the
\href{https://github.com/AzKurban-edX-DS/Capstone-MovieLens/blob/main/r/src/support-functions/common-helper.functions.R\#L59}{naive\_model\_RMSE}
one:

@@

\includegraphics{capstone-movielens-report.draft4_files/figure-latex/unnamed-chunk-43-1.pdf}

@@

\subsubsection{Taking into account User
Effect}\label{taking-into-account-user-effect}

@@

\includegraphics{capstone-movielens-report.draft4_files/figure-latex/unnamed-chunk-44-1.pdf}

@@

\includegraphics{capstone-movielens-report.draft4_files/figure-latex/unnamed-chunk-56-1.pdf}
